### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: s/fix-point/fixpoint/

In ML-like languages, datatypes are presented as a
\emph{sum-of-products}. A datatype is defined by a finite sum of
constructors, each composed by a product of arguments. To embrace
these datatypes, we need to capture this grammar.

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we define the syntax of a language. To give
it a semantics, we \emph{interpret} these codes back into the type
theory. Hence, codes act as labels, while the type theory provides the
computational machinery.

Codes being mere labels, we can inspect them, hence regaining
structural information. We can also compute over them: deriving new
codes from others, or even functions on them.

@@

To embed inductive types in our type theory, we define a universe of
\emph{descriptions} (Figure~\ref{fig:desc_universe}). As expected, the
code of this universe mimics the grammar of inductive types. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. Descriptions are terminated by $\DUnit$, containing no
payload. To make recursive arguments, we introduce the $\DIndx{\!}$
code. It interprets to a hole that will be closed by the fix-point. We
can also read $\V{D}$ as describing an endofunctor on $\Set$, with
$\descop{\V{D}}{\!}$ being the object part.

@@

\subsection{Fix-point}
\label{sec:desc-fix-point}

\begin{wstructure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fix-point
            <- Interpret D with (Mu D) as sub-objects
\end{wstructure}

So far, we have used our universe of descriptions to build pattern
functors. Being polynomial functors, they all admit a fix-point. We
construct the fix-point by \emph{tying the knot} of the
interpretation: the holes left open by the functor are filled by its
own recursive definition:

>>>>

In ML-like languages, datatypes are presented as a
\emph{sum-of-products}. A datatype is defined by a finite sum of
constructors, each composed by a product of arguments. To embrace
these datatypes, we have to capture this grammar.

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we define the syntax of a language. To give
it a semantics, we \emph{interpret} these codes back into the type
theory. Hence, codes act as labels, while the type theory provides the
computational machinery. Codes being mere labels, we can inspect them,
hence regaining structural information. We can also compute over them:
deriving new codes from others, or even functions on them.

@@

To embed inductive types in our type theory, we define a universe of
\emph{descriptions} (Figure~\ref{fig:desc_universe}). As expected, the
code of this universe mimics the grammar of inductive types. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. Descriptions are terminated by $\DUnit$, containing no
payload. To make recursive arguments, we introduce the $\DIndx{\!}$
code. It interprets to a hole that will be closed by the fixpoint. We
can also read $\V{D}$ as describing an endofunctor on $\Set$, with
$\descop{\V{D}}{\!}$ being the object part.

@@

\subsection{Fixpoint}
\label{sec:desc-fix-point}

\begin{wstructure}
<- Build the fixpoint of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as sub-objects
\end{wstructure}

So far, we have used our universe of descriptions to build pattern
functors. Being polynomial functors, they all admit a fixpoint. We
construct the fixpoint by \emph{tying the knot} of the
interpretation: the holes left open by the functor are filled by its
own recursive definition:

### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: clean-up some \note{}s.
Desc@ICFP: s/fix-point/fixpoint/

Let us consider the consequences. First, we discover that the type
theory does not need to be extended with the type former $\EnumU$, the
constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
above, by fix-point over the pattern functor denoted by
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
now replaced by codes and their fix-point in the universe of
descriptions. In essence, our next step is similar: we are going to
condense the entire naming scheme of datatypes \emph{in itself}.

@@

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of datatypes as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

\note{
Maybe a bit of perspective would be good: 
Everything stands up with the Mu.
Induction principle defined over Mu-things.
}

@@

\begin{wstructure}
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fix-point
\end{wstructure}

@@

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
the full-blown free monad construction.

>>>>

Let us consider the consequences. First, we discover that the type
theory does not need to be extended with the type former $\EnumU$, the
constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
above, by fixpoint over the pattern functor denoted by
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
now replaced by codes and their fixpoint in the universe of
descriptions. In essence, our next step is similar: we are going to
condense the entire naming scheme of datatypes \emph{in itself}.

@@

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fixpoint of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of datatypes as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

@@

\begin{wstructure}
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fixpoint
\end{wstructure}

@@

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fixpoint operation ties the knot and gives us
the full-blown free monad construction.

### 2010-04-01 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: clean-up some \note{}s.
Desc@ICFP: s/fix-point/fixpoint/

\note{ Need care: motivating motivation of indexing. }

@@

Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
\To \Set^{\V{I}}$.The fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

>>>>

@@

Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
\To \Set^{\V{I}}$.The fixpoint of inductive families is defined over
these indexed functors, hence computing a fixpoint of the entire
\emph{family} of functors:
